---
title: "lung-cancer-clustering"
output:
  md_document:
    variant: markdown_github
date: "2024-12-05"
---

```{r, results='hide'}
library(Seurat)
library(patchwork)
library(ggplot2)
library(dplyr)
library(RCurl)
library(AnnotationHub)
set.seed(30)
```
Adding metadata
```{r}
#filepaths <- names(assays)
meta <- read.delim("samplesheet.tsv")

#making a better metadata table
get.par <- function(string){return(substring(string,1,12))}
get.tumor <- function(string){return(substring(string,14))}

metaf <- meta %>% mutate("participant" = get.par(sample_id)) %>% mutate("tumor.type" = get.tumor(sample_id))

```


```{r}
assays <- list()

for(type in metaf$tumor.type){
  files <- metaf %>% dplyr::filter(tumor.type == type) %>% dplyr::select(filename)
  for(file in files[[1]]){
    data <- Read10X_h5(file)
    seurat.obj <- CreateSeuratObject(counts = data,
                                     min.features = 200,
                                     project = file)
    assign(file, seurat.obj)
  }
  assays[[type]] <- seurat.obj
}

merged.obj <- merge(x = assays$normal,
                      y = c(assays$edge, assays$middle, assays$core), 
                      add.cell.id = c("ctrl", "edge", "middle", "core"))

merged.obj <- JoinLayers(merged.obj)

head(merged.obj@meta.data)
#for(name in list.files("./counts", full.names = T)){
#  if(grepl("filtered.h5", name)){
#    assays[name] <- CreateSeuratObject(counts = #Read10X_h5(name), project = name)
    #}
#}

```
```{r}
#novelty score
# Add number of genes per UMI for each cell to metadata
merged.obj$log10GenesPerUMI <- log10(merged.obj$nFeature_RNA) / log10(merged.obj$nCount_RNA)

#add mitoRatio

# Compute percent mito ratio
merged.obj$mitoRatio <- PercentageFeatureSet(object = merged.obj, pattern = "^MT-")
merged.obj$mitoRatio <- merged.obj@meta.data$mitoRatio / 100
```

```{r}
#create meta datatable
metadata <- merged.obj@meta.data
metadata$cells <- rownames(metadata)

#sample column 
metadata <- metadata  %>% mutate(sample = case_when(
                                      grepl("core", cells) ~ "core",
                                      grepl("edge", cells) ~ "edge",
                                      grepl("ctrl", cells) ~ "ctrl",
                                      grepl("middle", cells) ~ "middle")) %>% 
                        dplyr::rename(nUMI = nCount_RNA, nGene = nFeature_RNA)

head(metadata)
```

It looks like we're getting far fewer cells than expected (4K, really?)
```{r}
metadata %>% 
  	ggplot(aes(x=sample, fill=sample)) + 
  	geom_bar() +
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells")
```

Looks alright, for some reason ctrl is much different. Sampling issues or celltype differences?
```{r}
#plotting
metadata %>% ggplot(aes(colour = sample, x=nUMI, fill = sample)) + 
  geom_density(alpha=0.4) +
  scale_x_log10() + 
  theme_classic() + 
  geom_vline(xintercept = 400)
```
We have a pretty low gene detection rate at a mean of ~300. Ctl sample seems much more consistent which may indicate a lesser diversity of cell types.
```{r}
metadata %>% ggplot(aes(colour = sample, x=nGene, fill = sample)) + 
  geom_density(alpha=0.4) +
  scale_x_log10() + 
  theme_classic()

mean(metadata$nGene)
median(metadata$nGene)
```
Evaluating complexity (nGene/nUMI). Some really interesting stuff here, it looks like the edge sample is bimodal, which might actually represent the spatial stratification of the sample. The control has a bump at 0.9 which is indicative of some heterogeneity. I feel like 0.8 might be too harsh a cut-off and we'd lose something interesting in the edge sample. 


```{r}
metadata %>% ggplot(aes(colour = sample, x=log10GenesPerUMI, fill = sample)) + 
  geom_density(alpha=0.4) +
  theme_classic() +
  geom_vline(xintercept = 0.76)
  
```
No mitochondrial 
```{r warning=FALSE}
metadata %>% ggplot(aes(colour = sample, x=mitoRatio, fill = sample)) + 
  geom_density(alpha=0.4) +
  theme_classic() +
  scale_x_log10() +
  geom_vline(xintercept = 0.2)
  
```

Mitochondrial expression seems to prevalent in edge tumor cells that are highly expressive. Since mito expression is very broad we probably shouldn't filter cells on this basis.
```{r message=FALSE, warning=FALSE}
metadata %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  	geom_point(size = 0.9) + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 500) +
  	geom_hline(yintercept = 250) +
  	facet_wrap(~sample) 
```

## Filtering
```{r}
#first lets throw the metadata on the object
merged.obj@meta.data <- metadata

filtered.seurat <- subset(x = merged.obj, 
                          subset= (nUMI >= 400) & 
                          (nGene >= 200) & 
                          (log10GenesPerUMI >= 0.76))

```

```{r}
# Extract counts
counts <- GetAssayData(object = filtered.seurat, layer = "counts")
nonzero <- counts > 0


# Sums all TRUE values and returns TRUE if more than 10 TRUE values per gene
keep_genes <- Matrix::rowSums(nonzero) >= 10

# Only keeping those genes expressed in more than 10 cells
filtered_counts <- counts[keep_genes, ]
filtered.seurat <- CreateSeuratObject(filtered_counts, meta.data = filtered.seurat@meta.data)
```

Checking after filtering
```{r message=FALSE, warning=FALSE}
filtered.seurat@meta.data %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=mitoRatio)) + 
  	geom_point(size = 0.9) + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 400) +
  	geom_hline(yintercept = 200) +
  	facet_wrap(~sample) 
  
```


```{r}
filtered.seurat@meta.data %>% 
  	ggplot(aes(x=sample, fill=sample)) + 
  	geom_bar() +
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells")
```

## Cell cycle analysis

```{r}

cc_file <- getURL("https://raw.githubusercontent.com/hbc/tinyatlas/master/cell_cycle/Homo_sapiens.csv") 
cell_cycle_genes <- read.csv(text = cc_file)
```

Getting annotations 
```{r}
ah <- AnnotationHub()
ahDb <- query(ah, 
              pattern = c("Homo sapiens", "EnsDb"), 
              ignore.case = TRUE)
id <- ahDb %>%
        mcols() %>%
        rownames() %>%
        tail(n = 1)

edb <- ah[[id]]

annotations <- genes(edb, 
                     return.type = "data.frame")

annotations <- annotations %>%
        dplyr::select(gene_id, gene_name, seq_name, gene_biotype, description)
```

Getting gene IDs map to cell cycle ensembl IDs
```{r}
# Get gene names for Ensembl IDs for each gene
cell_cycle_markers <- dplyr::left_join(cell_cycle_genes, annotations, by = c("geneID" = "gene_id"))

# Acquire the S phase genes
s_genes <- cell_cycle_markers %>%
        dplyr::filter(phase == "S") %>%
        pull("gene_name")
        
# Acquire the G2M phase genes        
g2m_genes <- cell_cycle_markers %>%
        dplyr::filter(phase == "G2/M") %>%
        pull("gene_name")
```

Cell cycle scoring
```{r}
seurat.phase <- NormalizeData(filtered.seurat)
seurat.phase <- CellCycleScoring(seurat.phase, 
                                 g2m.features = g2m_genes, 
                                 s.features = s_genes)
head(seurat.phase@meta.data)  
```
```{r}
# Identify the most variable genes
seurat.phase <- FindVariableFeatures(seurat.phase, 
                     selection.method = "vst",
                     nfeatures = 2000, 
                     verbose = FALSE)
		     
# Scale the counts
seurat.phase <- ScaleData(seurat.phase)
```

```{r}
# Identify the 15 most highly variable genes
ranked_variable_genes <- VariableFeatures(seurat.phase)
top_genes <- ranked_variable_genes[1:15]

# Plot the average expression and variance of these genes
# With labels to indicate which genes are in the top 15
p <- VariableFeaturePlot(seurat.phase)
LabelPoints(plot = p, points = top_genes, repel = TRUE)
```
```{r}
# Perform PCA
seurat.phase <- RunPCA(seurat.phase)

# Plot the PCA colored by cell cycle phase
DimPlot(seurat.phase,
        reduction = "pca",
        group.by= "Phase",
        split.by = "Phase")


DimPlot(seurat.phase,
        reduction = "pca",
        group.by= "Phase")
```
There is some variation with G2M and G1 in both PCs. However, since this is a stratified tumor dataset, we should expect some differentiation. There won't be any need to regress out phase information.



Splitting data for SCT normalization. Here we are just regressing gene length
```{r message=FALSE, warning=FALSE, results='hide'}
split.seurat <- SplitObject(seurat.phase, split.by = "sample")
splits.assays <- sapply(split.seurat, SCTransform, vst.flavor = "v2")

```

```{r}
splits.assays$ctrl@assays
splits.assays$core@assays
```
## Clustering

SMARCB1, a known tumor suppressor is a good indicator we're on the right track for one of the most variable genes.

```{r}
seurat.phase <- RunUMAP(seurat.phase, dims = 1:40)
UMAPPlot(seurat.phase)
```
We can see some tight clustering for specific subgroups emerging, however its difficult to see the differentiation. Let's try adjusting the parameters.
```{r}
seurat.phase <- RunUMAP(seurat.phase, dims = 1:40, min.dist = 0.80, n.neighbors = 30)
UMAPPlot(seurat.phase)
```
Here we can see the differences in the main cluster. "Edge" and "CTRL" samples seem to occupy a similar space at the lower end. We can also see clusters that are exclusive to the tumor samples. This dataset looks like it could benefit from integration.  

```{r}
## Don't run this during class
# Select the most variable features to use for integration
integ_features <- SelectIntegrationFeatures(object.list = splits.assays, 
                                            nfeatures = 2000)
## Don't run this during class
# Prepare the SCT list object for integration
splits.assays <- PrepSCTIntegration(object.list = splits.assays, 
                                   anchor.features = integ_features)

## Don't run this during class
# Find best buddies - can take a while to run
integ_anchors <- FindIntegrationAnchors(object.list = splits.assays, 
                                        normalization.method = "SCT", 
                                        anchor.features = integ_features)

## Don't run this during class
# Integrate across conditions
seurat_integrated <- IntegrateData(anchorset = integ_anchors, 
                                   normalization.method = "SCT")
```

```{r}
# Run PCA
seurat_integrated <- RunPCA(object = seurat_integrated)

# Plot PCA
PCAPlot(seurat_integrated,
        split.by = "sample")  
```

```{r}
seurat_integrated <- RunUMAP(seurat_integrated, dims = 1:40, min.dist = 0.75, n.neighbors = 30)
UMAPPlot(seurat_integrated)   
```
Now we'll check the clustering.
```{r}
seurat_integrated <- FindNeighbors(object = seurat_integrated, 
                                dims = 1:40)
```

```{r}
# Determine the clusters for various resolutions                                
seurat_integrated <- FindClusters(object = seurat_integrated,
                               resolution = c(0.05,0.1,0.2, 0.4))
```

```{r}

Idents(object = seurat_integrated) <- "integrated_snn_res.0.05"
DimPlot(seurat_integrated,
        reduction = "umap",
        label = TRUE,
        label.size = 6)

Idents(object = seurat_integrated) <- "integrated_snn_res.0.1"
DimPlot(seurat_integrated,
        reduction = "umap",
        label = TRUE,
        label.size = 6)


```

Using 0.4 makes sense here. Let's check a cell phase:

```{r}
# Explore whether clusters segregate by cell cycle phase
DimPlot(seurat_integrated,
        label = TRUE, 
        split.by = "Phase")  + NoLegend()
```
Each of our clusters appear to be present in every cell state. In fact, G2M seems to appear more segregated at lower cell counts.

Checking for uninteresting features.
```{r}
# Determine metrics to plot present in seurat_integrated@meta.data
metrics <-  c("nUMI", "nGene", "S.Score", "G2M.Score", "mitoRatio")

FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = metrics,
            pt.size = 0.4, 
            order = TRUE,
            min.cutoff = 'q10',
            label = TRUE)
```
The number of genes might be a defining factor for clusters to the right. 

Next we'll check cluster identity manually.

```{r}
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("SMARCB1"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

```{r}
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("CD14", "LYZ"), 
            order = TRUE, 
            label = TRUE)
```
Finding cluster gene identities 
```{r}
DefaultAssay(seurat_integrated) <- "RNA"

# Create function to get conserved markers for any given cluster
get_conserved <- function(cluster, seurat){
  FindConservedMarkers(seurat,
                       ident.1 = cluster,
                       grouping.var = "sample",
                       only.pos = TRUE)
}
seurat_integrated.joined <- JoinLayers(seurat_integrated)
clusters <-  sapply(c(0:7),get_conserved, seurat = seurat_integrated.joined)
# Iterate function across desired clusters
#conserved_markers <- purrr::map2_dfr(c(4,0,6,2), get_conserved, seurat = seurat_integrated)
```

Identifying Cell Type Clusters. Here are some intermediate results that I've found via looking up gene marker identities on PangaoDB. Next steps would be subclustering and comparing with a non-integrated dataset.
```{r}
cluster.df.anno <- DataFrame(Cluster = c(0:7), 
                ID = c("Myeloid","Alveolar","Plasma: T Cells?","?","Fibroblasts", "Plasma", "T Cells", "?"), 
                Gene.Markers = c("LYZ, C1QA","PGR, SFTPB","CD3G, PRDM1","?","SLIT2, CRISPLD2","JCHAIN, SMARCB1, IGKC","SIGLEC6, FOXP1","?"))
cluster.df.anno
```
